Iterative Unfolding Optimization with the Mean Squared Error Metric

### 2019-10-01 NOvA/numu/mean_squared_error_metric/mse_metric.tex
Work in progress.

%================================================================================
\begin{frame}{A Simple Alternative Metric: Mean Squared Error}
	Suppose $\theta$ is the a true parameter to be estimated, and $\hat{\theta}$ is an estimator of the parameter.
	The mean squared error (MSE) can be decomposed into a sum of variance and bias squared.
	\begin{eqnarray} \label{eq:2}
	MSE &=& E[(\hat{\theta}-\theta)^2]\nonumber\\
	&=& E[\hat{\theta}^2-2\hat{\theta}\theta+\theta^2] = E[\hat{\theta}^2] -2\theta E[\hat{\theta}]+\theta^2 \nonumber\\
	&=& (E[\hat{\theta}^2]-E[\hat{\theta}]^2)+(E[\hat{\theta}]^2-2\theta E[\hat{\theta}]+\theta^2) \nonumber\\
	&=& V[\hat{\theta}]+b^2
	\end{eqnarray}
	, where $b=E[\hat{\theta}]-\theta$ is the bias of the estimator.\\
	Equation~\ref{eq:2} is called bias-variance decomposition.
	\vfill
	This is a very common metric in statistics and machine learning as well.
\end{frame}

>>>>

%================================================================================
\begin{frame}{A Simple Alternative Metric: Mean Squared Error}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \footnotesize
			Suppose $\theta$ is the a true parameter to be estimated, and $\hat{\theta}$ is an estimator of the parameter.
			The mean squared error (MSE) can be decomposed into a sum of variance and bias squared.
			\begin{eqnarray} \label{eq:2}
			MSE &=& E[(\hat{\theta}-\theta)^2]\nonumber\\
			&=& E[\hat{\theta}^2-2\hat{\theta}\theta+\theta^2] = E[\hat{\theta}^2] -2\theta E[\hat{\theta}]+\theta^2 \nonumber\\
			&=& (E[\hat{\theta}^2]-E[\hat{\theta}]^2)+(E[\hat{\theta}]^2-2\theta E[\hat{\theta}]+\theta^2) \nonumber\\
			&=& V[\hat{\theta}]+b^2
			\end{eqnarray}
			, where $b=E[\hat{\theta}]-\theta$ is the bias of the estimator.\\
			Equation~\ref{eq:2} is called bias-variance decomposition.
			\vfill
			This is a very common metric in statistics and machine learning as well.
	\end{column}
    \begin{column}{.5\textwidth}
      \includegraphics[width=\textwidth]{figures/mse_illustration_iter20.pdf}
    \end{column}
  \end{columns}
\end{frame}

%================================================================================
\begin{frame}{Mean Squared Error in the Context of Unfolding}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \footnotesize
				Given a true histogram $\pmb{\mu}=(\mu_1,...,\mu_M)$, where $\mu_i$'s are true counts in the $i$-th bin, $i=1,...,M$, unfolding can be viewed as a procedure that outputs a vector of estimators for the bin counts $\pmb{\hat{\mu}}=(\hat{\mu}_1,...,\hat{\mu}_M)$.

				Denoting the mean squared error for the $i$-th bin $MSE_i$, MSE for the histogram can be defined as
				\begin{eqnarray}
				  MSE &=& \frac{1}{M}\sum_{i=1}^M MSE_i \nonumber\\
				  &=& \frac{1}{M}\sum_{i=1}^M V_{ii}+\hat{b}_i^2
				\end{eqnarray}
				, where $V_{ii}=cov[\hat{\mu}_i,\hat{\mu}_i]$ and $\hat{b}_i$ is an estimator of $E[\hat{\mu}_i]-\mu_i$.
    \end{column}
    \begin{column}{.5\textwidth}
      \footnotesize
			Very often one wants to estimate more accurately bins with smaller absolute statistical uncertainties. In this case, weighted MSE can be used:
			\begin{eqnarray}
			  weighted \quad MSE &=& \frac{1}{M}\sum_{i=1}^M \frac{MSE_i}{\hat{\mu}_i} \nonumber\\
			  &=& \frac{1}{M}\sum_{i=1}^M \frac{V_{ii}+\hat{b}_i^2}{\hat{\mu}_i}
			\end{eqnarray}
			In this study, a minimum bin count is required to satisfy Poisson statistics, and is compared to the result without count constraint.
		\end{column}
	\end{columns}
\end{frame}

%================================================================================
\begin{frame}{Accessing MSE with Toy Monte Carlo}
  \begin{enumerate}
    \item Take the true spectrum. Generate 100 pseudo experiments by fluctuating each bin by Poisson with the true bin count as the parameter, i.e., average value.
    \item Smear each experiment by applying the normalized migration matrix to the true spectrum.
      \begin{equation}
        \nu_i=\sum_j\left(\frac{A_{ij}}{\sum_i A_{ij}}\right)\mu_j
      \end{equation}
      , where $\nu_i$ is the reco spectrum and $\mu_j$ is the true spectrum.
    \item Unfold each spectrum by a certain number of iterations.
    \item For each iteration, calculate MSE with the 100 spectra. Find the number of iterations with a minimum MSE.
  \end{enumerate}
\end{frame}

%================================================================================
\begin{frame}{Why My First Attempt Did Not Work Out}
  \begin{columns}
    \begin{column}{.5\textwidth}
    \begin{itemize}
      \footnotesize
      \item My first try led to exceedingly large bias, washing out the variance term completely.
      \item All ND analyzers have been using the ``migration matrix'' obtained by \texttt{CAFAna} directly. My first attempt used the ``normalized'' migration matrix instead, resulting in large unfolded counts at bins with almost no counts in true space.
        \begin{equation*}
          B_{ij}=\frac{A_{ij}}{\sum_i A_{ij}}
        \end{equation*}
      \item I used the raw ``migration matrix'' in my second attempt and got the same results as others. Probably double counting somewhere.
    \end{itemize}
    \end{column}
    \begin{column}{.5\textwidth}
      \includegraphics[width=\textwidth]{figures/unfolded_old_new_comparison.eps}
    \end{column}
  \end{columns}
\end{frame}

%================================================================================
\begin{frame}{Results: Unweighted MSE}
  \centering
  \includegraphics[height=\textheight]{figures/mse_vs_iter.pdf}
\end{frame}

%================================================================================
\begin{frame}{Results: Weighted MSE}
  \begin{columns}
    \begin{column}{.5\textwidth}
      \includegraphics[width=\textwidth]{figures/mse_vs_iter_true_weighted_mincnt_0.pdf}
    \end{column}
    \begin{column}{.5\textwidth}
      \includegraphics[width=\textwidth]{figures/mse_vs_iter_true_weighted_mincnt_100.pdf}
    \end{column}
  \end{columns}
\end{frame}
